### 2014-09-06 overview.tex
-xz

\subsection{Preliminaries}
We first introduce some notation used throughout this paper. Intuitively, a trace is a sequence of events recorded during the observed run. An event, $e=<t, id, inst>$, is a concrete representation of the runtime execution of a static instruction $inst$, such that (i) $t$ refers to the thread that issues the event $e$ (denoted $t^e$); (ii) $id$ is a unique identifier of the event (unless spcified otherwise, we use the index of an event in the trace as its id); and (iii) $inst$ is a static three-address instruction (involving at most three operands, which modern compilers commonly support).

For this paper, we utilize the instructions listed in Table~\ref{Ta:syntax}. When the variable does not appear on the left hand of an equation, such as $y$ in $x.f=y$, it may refer to a variable, a constant or event object creation expression $new (...)$.  The $bop$ stands for the binary operator, which may refer to $+, -, *, /, \%, \wedge, \vee$ in the assignment, or refer to $<, >, =, \wedge, \vee$ in the branch. The target of the branch event is not important in our scope, therefore, we may abbreviate the branch instruction as the boolean expression afterwards. The listed instructions suffice to represent all trace events of interest. This is because a concrete finite execution trace can be reduced to a straight-line loop-free call-free path program  (argument passing of the method call is modeled as assignments and the virtual call resolution is modeled as branches). This standard form of simplification preserves all the data-race-related information contained in the original trace.

\begin{table}
	\begin{center}
		\begin{tabular}{rcl}
			\multicolumn{1}{l}{{\tt s} $::=$} & & \\
			{\tt y = x.f} & $|$ & {\bf (heapr)} \\ 
			{\tt x.f = y}  & $|$ & {\bf (heapw)} \\ %\ $|$\ {\tt x.f = $c$}\
			{\tt z = x $bop$ y}\  & $|$ & {\bf (assign)} \\ %$|$\ {\tt z = $c$} $|$ {\tt z = new()}
			{\tt if (x $bop$  y) goto ...} & $|$ &  {\bf (branch)} \\
			{\tt lock(l)}\ $|$\ {\tt unlock(l)}  & $|$& {\bf (sync)} \\
			{\tt fork(t)}\ $|$\ {\tt join(t)}$|$ {\tt begin(t)}\ $|$\ {\tt end(t)} &  & {\bf (thread)}
		\end{tabular}
	\end{center}
	\caption{\label{Ta:syntax}Instruction Types}
\end{table}

{\bf Trace Representation \ } In the following, we assume that trace $\tau$ is represented as $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$, where
\begin{itemize}
\item  $\Gamma$ refers to the set of threads involved in the trace, which include $t_1$, $t_2$, \dots, $t_n$;
\item   $\{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}$ denotes the event sequences produced by each thread;
\item  $O$ is a function that assigns  to every event an integer value that denotes its scheduling order: $e_i$ is scheduled earlier than $e_j$ iff $O(e_i)<O(e_j)$; and
\item  $\theta$ is a function that maps every live variable to a value. 
%For example, given an event $e$ executing the instruction $x=y+z$, the mapping recorded in the trace may be $\theta(x)=3$,  $\theta(y)=2$,  $\theta(z)=1$.  
\end{itemize}
To ensure the integrity of $\theta$ as well simplify downstream encoding steps performed by the analysis already at the syntactic level, we rewrite the raw input trace into SSA form, such that each variable is defined exactly once.

@@

%TODO this section is inconsistent with others. change it!
\subsection{Constraint System}
The predictive analysis takes the program and a trace as the input.  Suppose it starts with the following trace of the program in  Figure \ref{fig:running}:  $e_1 e_2 e_3 e_4 e_5$, where the line number is used as the event id (subscript). A potential race is between the pair, $(e_2, e_5)$, as they access the shared variable $x$ from different threads and one is a write.

Given the trace, $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$, the predictive analysis computes a new trace,  $\tau'=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O', \theta'>$. $\tau'$ preserves the same event sequence for each thread, i.e., $T_1$ still executes $e_1 e_2 e_3$ and $T_2$ still executes $e_4 e_5$, but it reschedules the events from different threads so that the potential race pair runs concurrently, i.e., $e_2$ and $e_5$ run concurrently. When the schedule changes, the variables may get different values. Therefore, we need to compute both the new schedule $O'$ and the new value mapping $\theta'$, in order to witness a race. To guarantee that $\tau'$ is a feasible trace of the program $P$, the computation needs to respect a set of constraints, which include the value constraints, control flow constraints and synchronization constraints~\footnote{We omit the synchronization constraints in this paper, which are already well explained in existing techniques~\cite{yannis, pecan}.}. We explain them briefly in the following. The full details are explained in Section~\ref{sec:relax1} and the feasibility guarantee is explained in Section~\ref{sec:guarantee}.

@@

{\bf Race Condition\ } The potential race pair involves two accesses of the same shared location from different threads, where at least one access is a write. Given any potential race pair, e.g., ($e_2$, $e_5$), it is a real race if and only if the two accesses can occur at the same time (race condition), which is captured by the race condition constraint: 
$O'(e_2) = O'(e_5)$.
The race condition constraint--- together with the other constraints --- guarantees the feasibility of the predicted race if a solution is found for the overall constraint system.

{\bf Intra-thread Constraints\ } The predictive analysis needs to preserve the same event sequence for each thread, which further requires

\begin{itemize}
\item {\bf Control Flow Constraints\ } Each thread takes the same control flows (or branch decisions) to reproduce the events.
 For the running example (ignoring the statements in red), this yields: $\theta'(R_{\tt y}^4) > 2 \equiv {\bf true}$.
That is, in the predicted trace $\tau'$, the value of variable {\tt y} read at $e_4$ should be greater than $2$.

\item {\bf Intra-thread Order Constraints\ } The events should follow the same thread-local order as in the original trace. This constraint is imposed by the fact that the two runs share the same instruction sequence and take the same control flows. 
We obtain the  formula: $O'(e_1) < O'(e_2) < O'(e_3) \wedge O'(e_4) < O'(e_5)$.
That is, in the predicted trace $\tau'$, the three events from $T_1$ are ordered in the same way as in $\tau$, and similarly, the branch event $e_4$ in $T_2$ still executes before the event $e_5$ inside the branch body.

\item {\bf Intra-thread Value Constraints\ } The events should respect the value constraints imposed by each instruction.
 For our running example, we obtain:
$$
\begin{array}{l}
	\theta'(W_{\tt x}^0) = 0 \wedge \theta'(W_{\tt y}^0) = 0 \wedge \theta'(W_{\tt y}^1) = 3 \\ 
	\wedge	\theta'(W_{\tt x}^2) = 1 \wedge \theta'(W_{\tt y}^3) = 5
	\end{array}
$$
As an example, $\theta'(W_{\tt y}^3)= 5$ denotes that the predicted run still assigns the value 5 to the variable {\tt y}  at event $e_3$.
\end{itemize}

@@

{\bf Inter-thread Value Constraints\ } The inter-thread value constraints capture what writes the read events read from and under what scheduling condition the value flow occurs.  The resulting formula for our example is
$$
\begin{array}{lcl}
	& & (\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^0) \wedge O'(e_4) < O'(e_1) ) \\
& \bigvee &
	(\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^1) \wedge O'(e_1) < O'(e_4) < O'(e_3)) \\
& \bigvee &
	(\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^3) \wedge O'(e_3) < O'(e_4))
\end{array}
$$    
Notice, importantly, that the formula associates the value constraints with the scheduling order constraints. 
 As an example, 
$\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^1) \wedge O'(e_1) < O'(e_4) < O'(e_3)$ means that, in the predicted run, the read $e_4$ reads from the write $e_1$, under the condition that $e_4$ happens after $e_1$ and no other  writes (such as $e_3$) interleave them.

The formuals from the different encoding steps are conjoined and sent to the off-the-shelf solver such as {\sf z3}.  The race under analysis is real if the solver returns a solution, which includes the scheduling order and the values for the variables.

{\bf Unexplored Branches\ } Beyond the encoding steps so far, which focus on the given trace, we can often encode constraints along unexplored branches. In our running example, this is essential to discover the race between lines {\tt 2} and {\tt 7}. We explore the unexecuted branch statically and collect constraints during the exploration. We then form a disjunction  of the original constraints and the collected constraints to include more possible  paths. The solver may pick up the unexplored branch to detect a race whenever it is necessary. Back to our example, the event $e_7$ is in the unexplored branch, our solver is able to identify the race between $e_7$ and $e_2$ after encoding the unexplored branch.

For the racy events, $e_2$ and $e_7$, we first specify the race condition constraint  $\theta'(R^4_y)\leq 2\rightarrow O'_{e_2}=O'(e_7)$. Note that the race condition constraint associates the race with the branch condition that enables each event.  Also we need to encode (1) the intra-thread constraints, such as the  order constraints $O'(e_4)<O'(e_5)\wedge O'(e_4)<O'(e_7)$, and (2) the inter-thread constraints, which remain the same as before. With these constraints, the solver confirms the race pair, ($e_2$, $e_7$). More details are explained in Section~\ref{sec:relax2}.

{\bf Highlights \ } In this example, our analysis finds two more pairs of races beyond existing analyses~\cite{yannis, jeff}: $(e_2, e_5)$ and $(e_2, e_7)$.
For the first race, we do not require $e_4$ to read from $e_3$ to retain the original value, as in approaches. Rather, we allow $e_4$ to read from $e_1$ while still preserving the feasibility of the branch. This relaxation allows the events after $e_4$ to run concurrently with the events before $e_3$, leading to the detection of the first race.  For the second race, we cover unexecuted branches and ask the solver to compute a feasible schedule that exercises them. The neighboring branches are likely to contain racy events if the executed branches contain racy events.

>>>>

\subsection{Preliminaries}
We first introduce some notation used throughout this paper. Intuitively, a 
trace is a sequence of events recorded during the observed run. An 
event, $e=<t, id, inst>$, is a concrete representation of the runtime 
execution of a static instruction $inst$, such that (i) $t$ refers to the 
thread; (ii) $id$ is a unique identifier of the event (unless specified 
otherwise, we use the index of an event in the trace as its id); 
and (iii) $inst$ is a static three-address instruction (involving at most three operands).

For this paper, we utilize the instructions listed in 
Table~\ref{Ta:syntax}. When the variable does not appear on the left hand 
side of an equation, such as $y$ in $x.f=y$, it may refer to a variable, a 
constant or an object creation expression $new (...)$.  Symbol $bop$ stands 
for a binary operator, which may refer to $+, -, *, /, \%, \wedge, \vee$ in 
an assignment, or  $<, >, =, \wedge, \vee$ in a branch instruction. 
The target of the branch instruction is not important in our scope, 
therefore, we may abbreviate a branch instruction as its boolean expression 
afterwards. The listed instructions suffice to represent all trace events 
of interest. 
Note that 
%This is because 
a concrete finite execution trace can be reduced to a straight-line 
loop-free call-free path program  (argument passing of a method call 
is modeled as assignments and virtual call resolutions are modeled as 
branches). 
%This standard form of simplification preserves all the data-race-related information contained in the original trace.

\begin{table}
	\begin{center}
		\begin{tabular}{rcl}
			\multicolumn{1}{l}{{\tt s} $::=$} & & \\
			{\tt y = x.f} & $|$ & {\bf (heapr)} \\ 
			{\tt x.f = y}  & $|$ & {\bf (heapw)} \\ %\ $|$\ {\tt x.f = $c$}\
			{\tt z = x $bop$ y}\  & $|$ & {\bf (assign)} \\ %$|$\ {\tt z = $c$} $|$ {\tt z = new()}
			{\tt if (x $bop$  y) goto ...} & $|$ &  {\bf (branch)} \\
			{\tt lock(l)}\ $|$\ {\tt unlock(l)}  & $|$& {\bf (sync)} \\
			{\tt fork(t)}\ $|$\ {\tt join(t)}$|$ {\tt begin(t)}\ $|$\ {\tt end(t)} &  & {\bf (thread)}
		\end{tabular}
	\end{center}
	\caption{\label{Ta:syntax}Instruction Types}
\vspace{-0.15in}
\end{table}

{\bf Trace Representation \ } A trace $\tau$ is represented as $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$, where
\begin{itemize}
\item  $\Gamma$ refers to the set of threads involved in the trace, which include $t_1$, $t_2$, \dots, $t_n$;
\item   $\{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}$ denotes the event sequences produced by each thread;
\item  $O$ is a function that assigns  to every event an integer value that denotes its scheduling order: $e_i$ is scheduled earlier than $e_j$ iff $O(e_i)<O(e_j)$; and
\item  $\theta$ is a function that maps every live variable to a value. 
%For example, given an event $e$ executing the instruction $x=y+z$, the mapping recorded in the trace may be $\theta(x)=3$,  $\theta(y)=2$,  $\theta(z)=1$.  
\end{itemize}
To ensure the integrity of $\theta$ as well simplify downstream encoding 
steps,
% performed by the analysis already at the syntactic level, 
we rewrite the raw input trace into SSA form, such that each variable is defined exactly once.

@@

%TODO this section is inconsistent with others. change it!
\subsection{Constraint System}
In this section, we are going to informally explain the two relaxations
we make which allow us to catch the two races missed by existing 
predictive analyses.
Our analysis takes the program and a trace as the input.  
Suppose it starts with the following trace of the program in  
Figure \ref{fig:running}:  $e_1 e_2 e_3 e_4 e_5$, where the line number 
is used as the event id (subscript).

\smallskip
\noindent
{\em I. Value Relaxation\ }
This allows us to catch the racy pair $(e_2, e_5)$.
%A potential race is between 
%the pair, $(e_2, e_5)$, as they access the shared variable $x$ from 
%different threads and one is a write. 
Given the trace, $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \},
 O, \theta>$, our analysis computes a new trace,  
$\tau'=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O', \theta'>$.
 $\tau'$ preserves the same event sequence for each thread, i.e., $T_1$ still 
executes $e_1 e_2 e_3$ and $T_2$ still executes $e_4 e_5$, but it reschedules 
the events from different threads so that the potential race pair runs 
concurrently, i.e., $e_2$ and $e_5$ run concurrently. When the schedule 
changes, the variables may get different values. Therefore, we need to 
compute both the new schedule $O'$ and the new value mapping $\theta'$, in 
order to witness a race. To guarantee that $\tau'$ is a feasible trace of 
the program $P$, the computation needs to respect a set of constraints, 
which include the value constraints, control flow constraints and 
synchronization constraints~\footnote{We omit the synchronization constraints 
in this paper, which are already well explained in existing 
techniques~\cite{yannis, pecan}.}. We explain the constraints 
briefly in the following. Details are in Section~\ref{sec:relax1} 
and the soundness guarantee is explained in Section~\ref{sec:guarantee}.

@@

{\bf Race Condition\ } A potential race pair involves two accesses of 
the same shared location from different threads, where at least one is write. 
Given any potential race pair, e.g., ($e_2$, $e_5$), it is a real race 
if and only if the two accesses can occur at the same time (race condition), 
which is captured by the race condition constraint: $O'(e_2) = O'(e_5)$.
The race condition constraint--- together with the other constraints --- 
guarantees the feasibility of the predicted race if a solution is found 
for the overall constraint system.

{\bf Intra-thread Constraints\ } Our analysis needs to preserve the 
same event sequence for each thread, which requires

\begin{itemize}
\item {\bf Control Flow Constraints\ } Each thread takes the same control flow
(or branch decisions) to reproduce the events.
 For the running example (ignoring the statements in red), this 
yields: $\theta'(R_{\tt y}^4) > 2 \equiv {\bf true}$.
That is, in the predicted trace $\tau'$, the value of variable {\tt y} 
read at $e_4$ should be greater than $2$.

\item {\bf Intra-thread Order Constraints\ } The events should follow 
the same thread-local order as in the original trace. This constraint 
is imposed by the fact that the two runs share the same instruction 
sequence and take the same control flow. 
We obtain the  formula: $O'(e_1) < O'(e_2) < O'(e_3) \wedge O'(e_4) < O'(e_5)$.
%That is, in the predicted trace $\tau'$, the three events from $T_1$ are 
%ordered in the same way as in $\tau$, and similarly, the branch 
%event $e_4$ in $T_2$ still executes before event $e_5$ inside the 
%branch body.

\item {\bf Intra-thread Value Constraints\ } The events should respect the 
constraints imposed by each instruction.
 For our running example, we obtain:
$$
\begin{array}{l}
	\theta'(W_{\tt x}^0) = 0 \wedge \theta'(W_{\tt y}^0) = 0 \wedge \theta'(W_{\tt y}^1) = 3 \\ 
	\wedge	\theta'(W_{\tt x}^2) = 1 \wedge \theta'(W_{\tt y}^3) = 5
	\end{array}
$$
As an example, $\theta'(W_{\tt y}^3)= 5$ denotes that the predicted run still assigns the value 5 to the variable {\tt y}  at event $e_3$.
\end{itemize}

@@

{\bf Inter-thread Value Constraints\ } These constraints capture what writes the read events read from and under what scheduling condition the value flow occurs.  The resulting formula for our example is
$$
\begin{array}{lcl}
	& & (\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^0) \wedge O'(e_4) < O'(e_1) ) \\
& \bigvee &
	(\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^1) \wedge O'(e_1) < O'(e_4) < O'(e_3)) \\
& \bigvee &
	(\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^3) \wedge O'(e_3) < O'(e_4))
\end{array}
$$    
Notice, importantly, that the formula associates the value constraints 
with the scheduling order constraints. 
As an example, 
$\theta'(R_{\tt y}^4)=\theta'(W_{\tt y}^1) \wedge O'(e_1) < O'(e_4) < O'(e_3)$ 
means that, in the predicted run, the read $e_4$ reads from the write $e_1$, 
under the condition that $e_4$ happens after $e_1$ and no other  
writes (such as $e_3$) interleave them.

The formulas from the different encoding steps are conjoined and sent to an 
off-the-shelf solver such as {\sf z3}.  The race under analysis is real 
if the solver returns a solution, which includes the scheduling order 
and the values for the variables.

\smallskip
\noindent
{\em II. Path Relaxation\ } 
%Next, we show how we catch the second missing race
%($e_2$, $e_7$) with another relaxation. 
Besides the first relaxation, which focuses on the given trace, our
technique also encodes constraints along unexplored branches. In our 
running example, this is essential to discovering the race between 
lines {\tt 2} and {\tt 7}. We explore the unexecuted branch statically 
and collect constraints during the exploration. We then form a disjunction  
of the original constraints and the collected constraints to include more 
possible  paths. The solver may pick up the unexplored branch to detect 
a race. Back to our example, event $e_7$ is in the unexplored branch, \tool
is able to identify the race between $e_7$ and $e_2$ after encoding 
the unexplored branch.

For the racy events, $e_2$ and $e_7$, we first specify the race condition 
constraint  $\theta'(R^4_y)\leq 2\rightarrow O'_{e_2}=O'(e_7)$. Note that 
the constraint associates the race with the branch condition that 
enables each event.  Also we need to encode (1) the intra-thread constraints, 
such as the  order constraints $O'(e_4)<O'(e_5)\wedge O'(e_4)<O'(e_7)$, 
and (2) the inter-thread constraints, which remain the same as before. 
With these constraints, the solver confirms the race pair, ($e_2$, $e_7$). 
More details are in Section~\ref{sec:relax2}.

\smallskip
\noindent
{\em Highlights \ } In this example, \tool finds two more pairs of races 
beyond existing analyses~\cite{yannis, jeff}: $(e_2, e_5)$ and $(e_2, e_7)$.
For the first race, we do not require $e_4$ to read from $e_3$ to 
retain the original value (i.e. the true branch outcome), as in the
existing approaches. Rather, we allow $e_4$ to read from $e_1$ while 
still preserving the feasibility of the path. This relaxation allows 
the events after $e_4$ to run concurrently with the events before 
$e_3$, leading to the detection of the first race.  For the second race, 
we cover unexecuted branches and query the solver to compute a feasible 
schedule that exercises them. The neighboring branches are likely to 
contain racy events if the executed branches contain racy events.
